---
title: "PLSR"
output: html_document
date: "2024-02-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 05

## PLS Model

---
title: "05CalibrationModels_PLS_Loop_ARC"
output: html_document
date: "2023-07-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\_ two ways of doing predictions (1) with an existing model that was built some time ago and (2) in this case when you have the spectral and wet chem data and you have to develop the model

Reference data = wet chemistry data of the full 100 samples (being used to develop the model)

QUESTION:

-   What does this portion of the code do?: spectraldata.fin\<-spectraldata[is.element(spectraldata\$SSN, df.f\$SSN),]

-   assuming the file spectral data is reading is referring to the selected samples from section 03?

-   or which method of splitting calibration and training data is used? KS or the random setseed123 used later in this script?

```{r}
library(dplyr)
library(prospectr)
library (globals)
library(stringr)
library(ggplot2)
library(here)

```

Load Reference data

```{r}

# spectraldata=read.csv("Raw_spectra-MIR.csv")
spec_trt <- readRDS("spec_trt_MIR.RDS")
spec_trt <- spec_trt[,-2] #removing the data for the first wavenumber since it's all NAN's


df.f<-readRDS("df.f.RDS")

threshold_na <- 0.85 #for a 85% cut-off

df.f<-
  df.f %>% select(where(~mean(is.na(.)) < threshold_na))

spectraldata.fin.fin<-spec_trt[is.element(spec_trt$SSN, df.f$SSN),] ## FUSI Still unclear what this is for. Changed to reference spec_trt
 
```

## Partial Least Squares Regression

Loading libraries

```{r, warning=FALSE}
##################### Random Forest Modelling #################################
library(randomForest)
library(caret)
library(pls)
library(data.table)
#If package ithir is not available for your version of R
#Use the 3 below lines to install package ithir
# install.packages("devtools") 
# library(devtools)
# install_bitbucket("brendo1001/ithir/pkg") or devtools::install_bitbucket("brendo1001/ithir/pkg")

library(ithir)
#used to extract model statistics like RMSE etc.
#using a function called: goof: see here: https://rdrr.io/rforge/ithir/src/R/goof.R 
```

### PLSR Model

```{r}

#Available properties to predict
#Incase you want to predict only selected properties,
#get property position by running line 66. Remove hash sign between ")" and "["
#symbols on line 67. Edit properties position and run line 67.
#Always ensure position 1 in always included.

#FUSI EDIT 
# changed the call to reference df.f and not df1 because df.f has columns removed for samples that don't have enough data 
# also removed the columns for char and char type

names(df.f)
slprptr<-names(df.f[-c(2:4)]) #FUSI EDIT: removing metadata columns

pred<-as.data.frame(spec_trt[,1])
colnames(pred)<-"SSN"

mdl.stats<-NULL#Model stats container

#FUSI EDIT: started at 3, instead of 2 because the first column in the char type
for(p in 23:length(slprptr)){

#Select properties to predict one at a time and remove NAs  
df.sel<-df.f %>% select(SSN, slprptr[p]) %>% na.omit


#Plot and print Biochar properties boxplots
boxplot(df.sel[,slprptr[p]], las=2, xlab = slprptr[p], ylab = "")
dir.create("Plots_Boxplots_PLSR_MIR")
png(paste0(getwd(),"/Plots_Boxplots_PLSR_MIR/",slprptr[p],".png"))
print(boxplot(df.sel[,slprptr[p]], las=2, xlab = slprptr[p], ylab = ""))
dev.off()

# #Split samples inside loop for variables with many NAs
# #Set splitting proportion for the calibration and validation data
set.seed(123)
pool=df.f[sample(nrow(df.sel), round(0.3*nrow(df.sel),0)), ]
pool<-pool[order(pool$SSN),]
poolid<-pool$SSN

#Get calibration and validation datasets
val_df<-pool
cal_df1 <-subset(df.sel, !(df.sel$SSN %in% val_df$SSN))

# threshold to exclude the extreme 5% values
#KARARI UPDATE; 95% loses too many samples, changed to 99% 

cal_df <-subset(cal_df1, cal_df1[,2]>quantile(cal_df1[,2], 0.01)&cal_df1[,2] <quantile(cal_df1[,2], 0.99))
val_df1 <-subset(df.sel, (df.sel$SSN %in% val_df$SSN))
val_df <-subset(val_df1, val_df1[,2]>quantile(val_df1[,2], 0.01)&val_df1[,2] <quantile(val_df1[,2], 0.99))
#FUSI EDIT: first chunk is initial code - for some reason wasn't actually orering the dataframe. second chunk is my edit
#renames the non-working version with suffice _or
val_df_or<-val_df[order(rownames(val_df)),]
cal_df_or<-cal_df[order(rownames(cal_df)),]

val_df<-setorder(val_df)
cal_df<-setorder(cal_df)

#Subset pre-treated spectra by available reference data
val_spec<-spec_trt[is.element(spec_trt$SSN, val_df$SSN),]
cal_spec<-spec_trt[is.element(spec_trt$SSN, cal_df$SSN),]
cal_spec<-cal_spec[order(cal_spec$SSN),]
val_spec<-val_spec[order(val_spec$SSN),]


#Get no of calibration and validation datasets
N_cal<-nrow(cal_spec)
N_val<-nrow(val_spec)

#Model data
Xcal.f=cal_spec[,-1]
Xval.f=val_spec[,-1]
dfcal.f=cal_df[,-1]
dfval.f=val_df[,-1]

###### PLSR SOC
maxc <- 10  ## number of max components
#FUSI EDIT: max was 25
pls.md <- plsr(dfcal.f ~ ., data = Xcal.f, ncomp=maxc, validation = "CV", segments = 10)#10-fold CV
#Each iteration of the cross-validation process involves training the model on 9 of the 10 segments and testing it on the remaining segment, repeating this process 10 times to ensure each segment is used as a test set once.

## plot RMSEP vs. number of components
plot(pls.md, "val", main=slprptr[p]) 

dir.create("Components_plots_PLSR_MIR")
png(paste0(getwd(),"/Components_plots_PLSR_MIR/",slprptr[p],".png"))
print(plot(pls.md, "val", main=slprptr[p]))
dev.off()

## no. components to use, the one with the smallest adj RMSEP value
RMSEP.obj<-RMSEP(pls.md)
str(RMSEP.obj)

RMSEP.obj$val[1:2,1,]
nc <- as.numeric(sub("comps", "", names(which.min(RMSEP.obj$val[1,1,2:dim(RMSEP.obj$val[1:2,1,])[2]]))))
nc

#Generate relevant model name
md.nm<-paste0("pls.md.", slprptr[p], ".nc", nc)

#Rename model with the looped Biochar property
assign(x = md.nm, value = get("pls.md", pos = .GlobalEnv), pos = .GlobalEnv)

## predict to validation dataset
pls.prd <- predict(pls.md, ncomp = nc, newdata = Xval.f)

## Return prediction statistics
val.stats=round(goof(dfval.f, pls.prd, type = "spec"),3)
val.stats<-bind_cols(Property=paste0(Property=slprptr[p],"_val"), Comps="", N=N_val, val.stats)
val.stats

## calibration statistics
pls.pc <- predict(pls.md, ncomp = nc, newdata = Xcal.f)

pls.cal=round(goof(dfcal.f, pls.pc, type = "spec"),3)
cal.stats<-bind_cols(Property=paste0(Property=slprptr[p],"_cal"), Comps=as.character(nc), N=N_cal, pls.cal)
cal.stats

################### Get model statistics #########################
mdstats<-bind_rows(cal.stats, val.stats)

#Create model stats labels for the plot
#FUSI EDIT: Removed comps from before N because plot was crowded should add back in and only remove in plot below 
slct.stats<-as.data.frame(t(mdstats[,c("Property","N","R2","RMSE","bias","RPIQ" )])) 
names(slct.stats)<-NULL
slct.stats<-bind_cols(rownames(slct.stats),slct.stats[,2])

#FUSI EDIT: added "comps" because it is in slct.stats 
valbls<-paste0(c("N","R2","RMSE","bias","RPIQ"), "\n")
valsts<-paste0(c(slct.stats[2,2],slct.stats[3,2],slct.stats[4,2],slct.stats[5,2],slct.stats[6,2]))
valstats<-paste(valbls,valsts)

#Bind all looped properties model stats
mdl.stats<-bind_rows(mdl.stats,mdstats)


lgth<-length(sort(dfval.f,decreasing=F))

seq.int(sort(dfval.f,decreasing=F)[1], sort(dfval.f,decreasing=F)[lgth],length.out=4)

#Plot validation plot
plot(dfval.f,pls.prd,pch=10,
     xlab=paste('Measured',names(val_df)[2],sep="_"),
     ylab=paste('Predicted',names(val_df)[2],sep="_"), 
     xlim = range(c(dfval.f,pls.prd)),
     ylim = range(c(dfval.f,pls.prd)),
     mtext(valstats[-1],side=3, at=c(seq.int(sort(dfval.f,decreasing=F)[1], sort(dfval.f,decreasing=F)[lgth],length.out=4)))
     )   ## plot the predicted vs. measured in the validation
abline(a = 0, b = 1)


dir.create("Plots_Validationplots_PLSR_MIR")
png(paste0(getwd(),"/Plots_Validationplots_PLSR_MIR/",slprptr[p],".png"))
print(plot(dfval.f,pls.prd,pch=10,
           xlab=paste('Measured',names(val_df)[2],sep="_"),
           ylab=paste('Predicted',names(val_df)[2],sep="_"), 
           xlim = range(c(dfval.f,pls.prd)),
           ylim = range(c(dfval.f,pls.prd)),
           mtext(valstats[-1],side=3, at=c(seq.int(sort(dfval.f,decreasing=F)[1], sort(dfval.f,decreasing=F)[lgth],length.out=4)))
           ))
abline(a = 0, b = 1)
dev.off()


################### Predict all samples #########################
prd.smpls <- predict(pls.md, spec_trt[,-1])

prd<-as.data.frame(prd.smpls)
df.prd<-bind_cols(SSN=rownames(prd),prd[,nc])
colnames(df.prd)<-c("SSN",slprptr[p])

pred<-merge(pred, df.prd, by="SSN", all.x = T)
}
#FUSI EDIT: End of for loop

```

```{r}


# Add two new columns based on whether the Property contains "_val" or "_cal"
mdl.stats_PLSR_MIR <- mdl.stats %>%
  mutate(Property_Name = gsub("(_cal|_val).*", "", Property),
         Data_Type = ifelse(grepl("_val", Property), "_val", "_cal"))

#Write model statistics and predicted values to the local drive
write.csv(mdl.stats_PLSR_MIR, paste0(getwd(),"/Model_Statistics_PLSR_MIR.csv"),row.names = F)
write.csv(pred, paste0(getwd(),"/Predicted_Biochar_Properties_PLSR_MIR.csv"),row.names = F)
getwd()

saveRDS(mdl.stats_PLSR_MIR,"mdl.stats_PLSR_MIR.RDS")
```

```{r}


mdl.stats_PLSR_MIR <- readRDS("mdl.stats_PLSR_MIR.RDS")

# Filter out rows with missing R2 values
mdl.stats_filtered <- mdl.stats_PLSR_MIR[complete.cases(mdl.stats_PLSR_MIR$R2), ]

# Order the properties based on R2 values in descending order for _cal data
ordered_props_cal <- mdl.stats_filtered[mdl.stats_filtered$Data_Type == "_cal", ]
ordered_props_cal <- ordered_props_cal[order(-ordered_props_cal$R2), "Property_Name"]
top_properties_cal <- head(ordered_props_cal, 10)

# Filter the original dataframe for the selected top properties and both _cal and _val data
mdl.stats_top <- mdl.stats_filtered[mdl.stats_filtered$Property_Name %in% top_properties_cal, ]

# Create a new factor variable to control the order of bars
mdl.stats_top$Property_Name <- factor(mdl.stats_top$Property_Name, levels = top_properties_cal)

# Plotting
cal_val_top_plot <- ggplot(mdl.stats_top, aes(x = Property_Name, y = R2, fill = Data_Type, group = Data_Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  labs(title = "Top Properties with Highest R2 for PLSR MIR",
       x = "Property",
       y = "R2") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels by 45 degrees

# Save the plot
ggsave(paste0(getwd(), "/cal_val_PLSR_MIR", ".png"),
       plot = cal_val_top_plot,
       width = 8,
       height = 6, 
       units = "in",
       dpi = 300)



# Extract relevant columns for the top properties and both _cal and _val data
mdl.stats_table <- mdl.stats_filtered[mdl.stats_filtered$Property_Name %in% top_properties_cal, c("Property_Name", "Data_Type","R2", "RPIQ", "RMSE", "bias", "N")]

# Create a new factor variable to control the order of rows in the table
mdl.stats_table$Property_Name <- factor(mdl.stats_table$Property_Name, levels = top_properties_cal)

# Order the rows based on Property_Name and Data_Type
mdl.stats_table <- mdl.stats_table[order(mdl.stats_table$Property_Name, mdl.stats_table$Data_Type), ]


# Create a table with the top properties, both _cal and _val data
mdl.stats_table_wide <- mdl.stats_table %>%
  pivot_wider(
    id_cols = c("Property_Name"),
    names_from = "Data_Type",
    values_from = c("R2","RPIQ", "RMSE", "bias", "N"),
    names_sep = "_"
  )

# Save the table to a CSV file
write.csv(mdl.stats_table, paste0(getwd(), "/Top_Properties_Stats_PLSR_MIR.csv"), row.names = FALSE)
```

## 
