---
title: "CEBUM"
output: html_document
date: "2023-12-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing Packages

```{r}

#if (!require("remotes")) install.packages("remotes")
#remotes::install_github("philipp-baumann/simplerspec")

#Info on package: 
  #https://github.com/philipp-baumann/simplerspec/ & 
  #https://github.com/philipp-baumann/simplerspec-read-filter-transform/blob/master/README.Rmd 

#install.packages(c("simplerspec","ggfortify"))
library(devtools)
#install_github("vqv/ggbiplot")
#library(simplerspec)
# Simplerspec is a universal file reader that allows one to read selected parameters instrument, optic and acquisition parameters)



suppressPackageStartupMessages({library(readxl)
                 library(plyr)
                 library(dplyr)
                 library(tidyr)
                 library(ggfortify)
                 library(tibble)
                 library(here)
                 library(ggbiplot)
                 library(tidyverse)
                 library(reshape)
                 library(reshape2)})
```

```{r}
files <- list.files(full.names = TRUE)
str(files)
```

The object `files` has the data structure *atomic vector*. *Atomic vectors* have six possible basic (*atomic*) vector types. These are *logical*, *integer*, *real*, *complex*, *string* (or *character*) and *mir*. Vector types can be returned by the R base function `typeof(x)`, which returns the type or internal storage mode an object `x`. For the `files` object it is

```{r}
# Check type of files object
typeof(files)
```

# Section 02

---
title: "02Spectra_Data_Cleaning_Exploration"
output: html_document
date: "2023-07-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### OBJECTIVE of SCRIPT

### Loading Libraries

```{r}
library(prospectr)
library(ggplot2)
library(dplyr)
library(reshape)
library(here)

#setwd("/Users/Biochars/Library/CloudStorage/OneDrive-CIFOR-ICRAF/Documents/0All_Training/Spectroscopy Data Training-17th May 2023/Data")
```

## Extracting spectra, removing meta data, and briefly visualizing

NOTES

-   Melt functionally takes a dataframe from wide to long format

```{r}

flnm<-"Raw_spectra-MIR.csv"

# Read spectra
raw <- read.csv(flnm)
raw[1:2,1:20]
names(raw)[1:20]
dim(raw)

############ Remove spectra metadata ######
 
raw0 <- raw[-c(1:36),-c(2:18)] #FUSI EDIT: removed first wavenumber because it was all NA's
## since the MIR samples included some blanks up to row 37
raw0[1:4,1:4]

wavenumbers <- round(as.numeric(substr(colnames(raw0[,-1]),2,100)),1)
##FUSI: this is another way to remove the X that appears in the column names since the values are numeric

colnames(raw0) <- c("SSN", wavenumbers)

raw0[1:5,c(1:4,1700:1704)]

# Create temporary SSNs incase there are repeat scans. The flat csv file retains the original assigned SSNs whether with repeats or not.

length(unique(raw0$SSN))

relabel <- length(grep('TRUE',duplicated(raw0[,1])))

ifelse(relabel>0, raw0[,1] <- paste(raw0$SSN, 1:nrow(raw0),sep = "-"),raw0[,1] <- raw0[,1])

#adds - 1, 2, 3 ... to id so each sample is unique

spec.m <- melt(as.data.frame(raw0), id="SSN")


p <- ggplot(data = spec.m, aes(x = as.numeric(as.vector(variable)),y = value,group = SSN)) +
  
  geom_line(size = 0.5, alpha = 0.1, col = 'brown') +
  
  ggtitle(strsplit(flnm,".csv")[[1]][1]) +
  
  xlim(rev(range(as.numeric(as.vector(spec.m$variable))))) +
  
  #note that the variable column refers to the wavenumber and the value column to the absorbance
  ylim(range(spec.m$value)) + 
  
  # ylim(c(0,1.3)) +
  
  xlab(expression("Wavenumbers (cm)"^-1)) +
  
  ylab("Aborbance units") + 
  #theme with white background
  theme_bw() +
  #eliminates background, gridlines, and chart border
  theme(
    plot.background = element_blank()
    ,panel.grid.major = element_blank()
    ,panel.grid.minor = element_blank()
  )
p <- p + theme(plot.title = element_text(hjust = 0.5))

p <- p + theme(legend.position = "none")

p <- p + theme(panel.background = element_rect(fill = "white"))

p

ggsave("Raw_spectra_plot_MIR.png")
################################################################################
```

### Cleaning Spectra

NOTES

-   The 'up' and 'down' cleanups are based on:
    -   So thinking about outliers: this is initially scanning the region between 2450-2500 - mainly this is probably a visual scan of the extracted dataset
    -   then maybe one identifies a peak with wavenumber 2498.2 and decides to remove this because it has an absorbance above 1.8
    -   the same is done for the peak at 3696.4 with absorbance below 1.1
    -   then the spectra are recombined

```{r}
################## Clean Spectra ##############

vrbs<-as.numeric(as.vector(spec.m$variable))

##### Up
spec.m[which(vrbs > 2450 & vrbs < 2500),]
bdup<-which(spec.m$variable == 2498.2 & spec.m$value > 1.8)

##### Down
spec.m[which(vrbs > 3600 & vrbs < 3700),]
bddwn<-which(spec.m$variable == 3696.4 & spec.m$value < 1.1)


#extracting only the rows/samples in the spec.m dataframe that meet the above criteria defined for the bdup and bddwn dataframes
spec.m2<-spec.m[which(spec.m$SSN %in% spec.m[c(bdup,bddwn),1]),]



bd <- ggplot(data = spec.m2, aes(x = as.numeric(as.vector(variable)),y = value,group = SSN)) +
  
  geom_line(size = 0.5, alpha = 0.1, col = 'brown') + 
  
  xlim(rev(range(as.numeric(as.vector(spec.m2$variable))))) +
  
  ylim(range(spec.m2$value)) + xlab(expression("Wavenumbers (cm)"^-1)) +
  
  ylab("Aborbance units") + theme_bw()

bd

#Now redefining the spec.m dataframe to omit the samples that meet the bdup and bddwn criteria as these are defined here as outliers

spec.m <- spec.m[which(!spec.m$SSN %in% spec.m[c(bdup,bddwn),1]),]

  
  
############## Average spectra if in duplicate

##FUSI edit 

# Remove the suffix from SSN
spec.m$SSN <- gsub("-\\d+$", "", spec.m$SSN)
spec.m$SSN <- as.factor (spec.m$SSN)

# Average the replicates
avg_spec <- spec.m %>% 
  group_by(SSN, variable) %>%
  summarise(value = mean(value, na.rm = TRUE))


# Cast the melted data back to wide format
raw0 <- dcast(avg_spec, SSN ~ variable, value.var = "value")

#previous version
#raw0 <- cast(spec.m, SSN~variable)
dim(raw0)

rownames(raw0) <- raw0$SSN

names(raw0) <- c("SSN", paste0("X", names(raw0)[-1]))

write.csv(raw0, paste0(getwd(),"/","model_",flnm),row.names = F)
```

```         
```

### Spectral Preprocessing

-   The first chunk is only necessary if you have a model spectra already, and you want to combine it with the new data so you can preprocess it together, you (likely) won't have this in the beginning i.e. if you are working on developing the model so you can skip this chunk

QUESTIONS

-   The following code seems to be doing the same thing as the 04Overlaying_spectra script - are there main differences?

TO DO

-   Read about the mean-centering function and why it's done here

Notes from Karari -

-   may have to look into database to see how to beef up dataset

-   if biochar does not share similar characteristics to Biochars data used to build ICRAF models ... ):

```{r}
#The following chunk is only necessary if you have a model spectra already, and you want to combine it with the 
# new data so you can preprocess it together, you (likely) won't have this in the beginning i.e. if you are working on developing the model 
# so you can skip this chunk 

####################### Read model data and preprocess #########################
# master<- read.csv("model-raw_spectra.csv")[,-c(2:17)]
# dim(master)
# master[1:5,1:5]
# 
# wavenumbers <- round(as.numeric(substr(colnames(master[,-1]),2,100)),1)
# 
# colnames(master) <- c("SSN", wavenumbers)
# 
# #Bind model and new data
# nwraw0 <- bind_rows(master,nwraw0)
# length(unique(nwraw0$SSN))
################################################################################



nwraw0 <- raw0

wavenumbers <- round(as.numeric(substr(colnames(nwraw0[,-1]),2,100)),1)
##FUSI: this is another way to remove the X that appears in the column names since the values are numeric

colnames(nwraw0) <- c("SSN", wavenumbers)

#FUSI EDIT: also removed bands below 650 as per Lago

##################################### Spectra pre-treatment ###################
#remove the CO2 region from all data

#FUSI EDIT: commenting out this section because for this dataset, co2 is empty 

# wavenumbers_1 <- colnames(nwraw0[,-1])
# co2 <- which(wavenumbers_1 < 2380 & wavenumbers_1 > 2350) # Get co2 bands
# nwraw0 <- nwraw0[,-c(co2+1)]
# dim(nwraw0)
# nwraw0[1:5,1:5]
```

### Savitzky Golay and Mean Centering

```{r}

#SavitzkyGolay pretreatment
#FUSI NOTES: Takes the first derivative and smooths the spectra

#often used as a preliminary preprocessing step to resolve overlapping signals, enhance signal properties, and suppress unwanted spectral features that arise due to nonideal instrument and sample properties."



spec_der<-as.data.frame(savitzkyGolay(nwraw0[,-1], w=17, p=2, m=1))

#Mean-centering function
center_colmeans <- function(x) {
  xcenter = colMeans(x)
  x - rep(xcenter, rep.int(nrow(x), ncol(x)))
}

#Mean-centering
spec_der.mc_trt<-center_colmeans(spec_der)

spec_trt<-bind_cols(SSN=nwraw0$SSN,spec_der.mc_trt)

#FUSI EDIT

saveRDS(spec_trt,"spec_trt_MIR.RDS")
 
```

### Plot of pretreted spectral (w/ Outliers)

```{r}

spec_pretreat <- melt(as.data.frame(spec_trt), id="SSN")

plot_pretreat <- ggplot(data = spec_pretreat, aes(x = variable,y = value,group = SSN)) +
  #as.numeric(as.vector(
  geom_line(size = 0.5, alpha = 0.1, col = 'brown') +
  
  ggtitle(strsplit(flnm,".csv")[[1]][1]) +
  
 # xlim(rev(range(as.numeric(as.vector(spec_trt$variable))))) +
  
  #note that the variable column refers to the wavenumber and the value column to the absorbance
 # ylim(range(spec_trt$value)) + 
  
  # ylim(c(0,1.3)) +
  
  xlab(expression("Wavenumbers (cm)"^-1)) +
  
  ylab("Mean-Centered Derivative of Aborbance") + 
  #theme with white background
  theme_bw() +
  #eliminates background, gridlines, and chart border
  theme(
    plot.background = element_blank()
    ,panel.grid.major = element_blank()
    ,panel.grid.minor = element_blank()
  )
plot_pretreat
```

### PCA Plots

QUESTIONS:

-   This is part of the overlay process to see if the model data casts in the same spectral space as the second dataset, correct?
-   prcomp function
    -   default centers the data - but we've already done mean-centering - so does this do nothing?

#### PCA with Outliers

```{r}

#Plot PCA scores plot overlay
dsnv <- spec_trt

pcs <- prcomp(dsnv[,-c(1:2)]) 
#FUSI EDIT: also removed the first data row since it's all NA's 

pcss <- pcs$x[,1:10]

pcss[1:6,]

plot(pcss)

#FUSI EDIT
#visualizing the loadings 
pcs$rotation
pcs

#FUSI EDIT: The original code was calib <- dim(master)[1]; but this refers to the master which as defined above is based on data used to develop the model
# I changed it to refer to nwraw0 here because that is the data being used to develop the model
#noting also that 'calib' has no significance to calibration here - just kept the same variable name for ease
calib <- dim(nwraw0)[1]


#calib <- dim(master)[1]

points(pcss[1:calib,1:2], col = "red")

points(pcss[-c(1:calib),1:2], col = "blue")

var <- round(summary(pcs)$importance[2,] * 100, 1)

scores <- cbind("Calib",as.data.frame(pcs$x[,1:5])) # get first 5 principal components

names(scores) <- c("set", colnames(scores[,-1]))

scores <- as.matrix(scores)

scores[-c(1:calib),1] <- "New samples"

scores <- as.data.frame(scores)

write.csv(scores, file = "./Calib and Pred scores MIR.csv", row.names = FALSE)

scores <- read.csv("./Calib and Pred scores MIR.csv")

#sp <- sp +  labs(color = "set")
sp <- ggplot(scores, aes(x = PC1, y =PC2, colour = set)) +
  
  geom_point(size = 0.8, alpha = 0.85 ) +
  
  ggtitle("PCA scores plot") +
  
  
  xlab(paste0("PC1 explains ", var[1], "% total variance")) +
  
  ylab(paste0("PC2 explains ", var[2], "% total variance")) +
  
  theme_bw() +
  
  theme(
    plot.background = element_blank()
    ,panel.grid.major = element_blank()
    ,panel.grid.minor = element_blank()
  )
sp <- sp + theme(plot.title = element_text(hjust = 0.5))

sp <- sp + scale_color_manual(values =c("brown","orange"))

sp

ggsave(filename  = "./Biochar_scores_MIR.png", height = 6, width = 6,sp)
  
```

## Fusi Edits

### PC 3 & 4

```{r}

sp <- ggplot(scores, aes(x = PC3, y =PC4, colour = set)) +
  
  geom_point(size = 0.8, alpha = 0.85 ) +
  
  ggtitle("PCA scores plot") +
  
  
  xlab(paste0("PC3 explains ", var[3], "% total variance")) +
  
  ylab(paste0("PC4 explains ", var[4], "% total variance")) +
  
  theme_bw() +
  
  theme(
    plot.background = element_blank()
    ,panel.grid.major = element_blank()
    ,panel.grid.minor = element_blank()
  )
sp <- sp + theme(plot.title = element_text(hjust = 0.5))

sp <- sp + scale_color_manual(values =c("brown","orange"))

sp

```

Biplot: the arrows represent the original variables, and the direction and length of the arrows indicate the variable's contribution to the principal components

-   the position of a point represents the projection of the observation into the reduced-dimensional space defined by the principal components

### Identifying and removing Outliers (and oxidized biochars) based on PCA

```{r}
SSN <- dsnv$SSN

# Combining sample IDs with principal component scores
scores_with_ids <- cbind(SSN, scores)

nine_outliers <- scores_with_ids  %>% 
  filter(PC2 <= -0.10)
 
#removing the samples that are outliers
spec_trt <- anti_join(spec_trt, nine_outliers, by = "SSN")


# Define the list of SSNs to remove
ssns_to_remove <- c("WA078289", "WA078290", "WA078291", "WA078292", "WA078293",
                    "WA078294", "WA078295", "WA078296", "WA078297", "WA078298",
                    "WA078299", "WA078300", "WA078301", "WA078302", "WA078303",
                    "WA078304", "WA078305", "WA078306")

# Remove rows with specified SSNs
spec_trt <- spec_trt %>%
  filter(!SSN %in% ssns_to_remove)

#resaving the spec_trt data without the outliers
saveRDS(spec_trt,"spec_trt_MIR.RDS")
```

#### PCA Without Outliers

```{r}

spec_trt <- readRDS("spec_trt_MIR.RDS")
#Plot PCA scores plot overlay
dsnv <- spec_trt

pcs <- prcomp(dsnv[,-c(1:2)]) 
#FUSI EDIT: also removed the first data row since it's all NA's 

pcss <- pcs$x[,1:10]

pcss[1:6,]

plot(pcss)

#FUSI EDIT
#visualizing the loadings 
pcs$rotation
pcs

#FUSI EDIT: The original code was calib <- dim(master)[1]; but this refers to the master which as defined above is based on data used to develop the model
# I changed it to refer to nwraw0 here because that is the data being used to develop the model
#noting also that 'calib' has no significance to calibration here - just kept the same variable name for ease
#Changed to reference spec_trt instead of dim(nraw0) because I removed outliers 
calib <- dim(spec_trt)[1]


#calib <- dim(master)[1]

points(pcss[1:calib,1:2], col = "red")

points(pcss[-c(1:calib),1:2], col = "blue")

var <- round(summary(pcs)$importance[2,] * 100, 1)

scores <- cbind("Calib",as.data.frame(pcs$x[,1:5])) # get first 5 principal components

names(scores) <- c("set", colnames(scores[,-1]))

scores <- as.matrix(scores)

scores[-c(1:calib),1] <- "New samples"

scores <- as.data.frame(scores)

write.csv(scores, file = "./Calib and Pred scores MIR.csv", row.names = FALSE)

scores <- read.csv("./Calib and Pred scores MIR.csv")

#sp <- sp +  labs(color = "set")
sp <- ggplot(scores, aes(x = PC1, y =PC2, colour = set)) +
  
  geom_point(size = 0.8, alpha = 0.85 ) +
  
  ggtitle("PCA scores plot") +
  
  
  xlab(paste0("PC1 explains ", var[1], "% total variance")) +
  
  ylab(paste0("PC2 explains ", var[2], "% total variance")) +
  
  theme_bw() +
  
  theme(
    plot.background = element_blank()
    ,panel.grid.major = element_blank()
    ,panel.grid.minor = element_blank()
  )
sp <- sp + theme(plot.title = element_text(hjust = 0.5))

sp <- sp + scale_color_manual(values =c("brown","orange"))

sp

ggsave(filename  = "./Biochar_scores_MIR.png", height = 6, width = 6,sp)
  
```

```{r}

SSN <- spec_trt$SSN

# Combining sample IDs with principal component scores
scores_with_ids <- cbind(SSN, scores)

nine_outliers <- scores_with_ids  %>% 
  filter(PC2 <= -0.10)

other_outliers <- scores_with_ids  %>% 
  filter(PC2 <= -0.060)
```

```{r}

# Calculate the absolute values of loadings
absolute_loadings <- abs(pcs$rotation)
#This line calculates the absolute values of loadings, where pcs is the result of the PCA. Loadings represent the contribution of each original variable to the principal components.

# Create a variable importance plot for the first few principal components
barplot(absolute_loadings[, 1:2], beside = TRUE, col = rainbow(5), 
        main = "Variable Importance for PC1-PC2", 
        #names.arg = colnames(absolute_loadings[, 1:2]), 
        cex.names = 0.7, las = 2)
#This line creates a bar plot representing the variable importance for the first two principal components (PC1 and PC2).

# Identify the top 15 loaded variables for each principal component
top_loaded_variables <- apply(absolute_loadings, 2, function(x) head(order(-x), 15))


# Extract loadings for the first two principal components
loadings_pc1 <- absolute_loadings[, 1]
loadings_pc2 <- absolute_loadings[, 2]

# Identify the top 15 loaded variables for each principal component
top_loaded_variables_pc1 <- head(names(sort(-abs(loadings_pc1))), 15)
top_loaded_variables_pc2 <- head(names(sort(-abs(loadings_pc2))), 15)

# Create a data frame to store the top loaded variables and their loadings
top_loaded_df <- data.frame(PC = c(rep(1, 15), rep(2, 15)),
                            Variable = c(top_loaded_variables_pc1, top_loaded_variables_pc2),
                            Loading = c(loadings_pc1[top_loaded_variables_pc1], loadings_pc2[top_loaded_variables_pc2]))

# Save the table to a CSV file
write.csv(top_loaded_df, file = "top_loaded_variables_pc1_pc2.csv", row.names = FALSE)

```

#### Plot of Pretreated Spectra w.o Outliers

```{r}

spec_pretreat <- melt(as.data.frame(spec_trt), id="SSN")

plot_pretreat <- ggplot(data = spec_pretreat, aes(x = variable, y = value, group = SSN, color = SSN)) +
  geom_line(size = 0.5, alpha = 0.1) + #,col='brown' # Set the line properties
  
  #ggtitle(strsplit(flnm,".csv")[[1]][1]) + 
  ggtitle("Pretreated Spectra with Some Outliers Removed") +  # Set the plot title
   # xlim(rev(range(as.numeric(as.vector(spec_trt$variable))))) +
  #note that the variable column refers to the wavenumber and the value column to the absorbance
  # ylim(range(spec_trt$value)) + 
  
  # ylim(c(0,1.3)) +

  xlab(expression("Wavenumbers (cm)"^-1)) +  # Set the x-axis label

  ylab("Mean-Centered Derivative of Absorbance") +  # Set the y-axis label

  theme_bw() +  # Set theme to black and white

  theme(
    plot.background = element_blank(),  # Remove plot background
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()   # Remove minor gridlines
  ) +
  
  scale_color_manual(values = c("WA078345" = "red", "WA078346" = "red", "WA078347" = "red", 
                                 "WA078348" = "red", "WA078349" = "red", "WA078350" = "red", 
                                 "Other" = "black"), 
                     guide = "none")  # Set manual color scale and remove the legend

print(plot_pretreat)

```

# Section 03

-   note:

    -   section 03: Kennard Stone is not used for the Cornell MIR files, because we don't need to split and determine which samples need to be sent for wet chem analysis since we already have the wet chem data (for whichever samples)

# Section 04

-   note:

    -   section 04: Overlaying, also not used here since we don't have any sort of reference model file

# Getting Reference Data

---
title: "05CalibrationModels_PLS_Loop_ARC"
output: html_document
date: "2023-07-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\_ two ways of doing predictions (1) with an existing model that was built some time ago and (2) in this case when you have the spectral and wet chem data and you have to develop the model

Reference data = wet chemistry data of the full 100 samples (being used to develop the model)

QUESTION:

-   What does this portion of the code do?: spectraldata.fin\<-spectraldata[is.element(spectraldata\$SSN, df.f\$SSN),]

-   assuming the file spectral data is reading is referring to the selected samples from section 03?

-   or which method of splitting calibration and training data is used? KS or the random setseed123 used later in this script?

```{r}
library(dplyr)
library(prospectr)
library (globals)
library(stringr)
library(ggplot2)
library(here)
```

Load Reference data

```{r}

# spectraldata=read.csv("Raw_spectra-MIR.csv")
spec_trt <- readRDS("spec_trt_MIR.RDS")
spec_trt <- spec_trt[,-2] #removing the data for the first wavenumber since it's all NAN's

# #rownames(spectraldata)<-spectraldata$SSN
# 
# 
# #Remove metadata columns when using raw file
#  names(spectraldata)
#  spectraldata<-spectraldata[-c(1:36),] #removing the standards
#  spectraldata<-spectraldata[,-2:-17] #removing metadata
#  
# # #Ensure all the wavenumbers are rounded to one decimal place
#  names(spectraldata)=round(as.numeric(str_sub(names(spectraldata),2)),1)
#  names(spectraldata)[1] <- "SSN"

#Read reference data
df1=read.csv("CEBUM_Reference_Data.csv")#, skip=5)#[,c(1,8:29)]

#TEMPORARY, UNTIL I RECIEVE THE REMAINING DATA 
#df1<-df1[c(1:84),-c(5:24)]


colnames(df1)[1]<-"SSN"
#colnames(df1)[3]<-"char_type"

#FUSI EDIT: put this cleanup chunk earlier to then filter out the columns where there isn't enough data, and then be able to define  slprptr based on the cleaned data 
# 

# #Remove unwanted characters in the data
# df.f[sapply(df.f, grepl, pattern = "<")] <- "NA"
dnmrc <- df1[,-c(1:9)] %>% mutate_if(is.character, as.numeric)
df1 <- bind_cols(df1[,c(1:9)],dnmrc)

#FUSI EDIT - errors out becauseo of duplicated? REVISIT
#Revisited - added back after selecting only the samples with ref data 
#KARARI

rownames(df1)<-df1$SSN


# 
# 
# ##################################### Spectra pre-treatment ###################
# #remove the CO2 region from all data
# co2rem<-which((as.numeric(colnames(spectraldata[-1]))) < 2380 & (as.numeric(colnames(spectraldata[-1]))) > 2350)
# spectraldata.f<-spectraldata[,-co2rem]
# 
# #Mean-centering function
# center_colmeans <- function(x) {
#   xcenter = colMeans(x)
#   x - rep(xcenter, rep.int(nrow(x), ncol(x)))
# }
# 
# #SavitzkyGolay pretreatment
# spec_der<-as.data.frame(savitzkyGolay(spectraldata.f[,-1], w=17, p=2, m=1))
# 
# #Mean-centering
# spec_der.mc_trt<-center_colmeans(spec_der)
# 
# spec_trt<-bind_cols(SSN=spectraldata$SSN,spec_der.mc_trt)

###############################################################################

#FUSI EDIT: ... define what raw0 is
#Also: this variable isn't used really
#spectraldata<- raw0


 #select only the common Biochar samples in spectral and wet chemie data
df.f<-df1[is.element(df1$SSN, spec_trt$SSN),]

#Create a df.f that isn't overided
df.f_master <-df.f

#FUSI EDIT: put this cleanup chunk earlier to then filter out the columns where there isn't enough data, and then be able to define  slprptr based on the cleaned data 
# 
# # #Remove unwanted characters in the data
# df.f[sapply(df.f, grepl, pattern = "<")] <- "NA"
# dnmrc <- df.f[,-1] %>% mutate_if(is.character, as.numeric)
# df.f <- bind_cols(SSN = df.f[,1],dnmrc)

#FUSI EDIT: 
#filtering and removing the columns (reference variables) where more than 70% of data is missing/NA

saveRDS(df.f,"df.f.RDS")
```

## Summary statistics

```{r}

library(skimr)
# Generate summary statistics
summary_stats <- skim(df.f[,-c(1:7)])


# Remove "numeric." prefix from column names
colnames(summary_stats) <- sub("numeric\\.", "", colnames(summary_stats))

# Remove some columns
summary_stats <- subset(summary_stats, select = -c(hist,skim_type))



# Write summary statistics to a CSV file
write.csv(summary_stats, file = "summary_statistics.csv")

```
